---
title: "SDS291 Final Project"
author: "Chaira Harder, Isabelle Elder, Sirohi Kumar"
format: html
editor: visual
---

```{r}

library(dplyr)
library(ggplot2)
library(tidyverse)
library(tidytuesdayR)

```

## INTRODUCTION

## METHODS

### Load the data:

```{r}

tuesdata <- tidytuesdayR::tt_load('2024-03-26')

team_results <- tuesdata$`team-results`
public_picks <- tuesdata$`public-picks`

```

### Data wrangling:

Before we use our data, we need to format some of the numbers by removing the % and convert the <chr> percentages to their numeric values. In our wrangling process, we will also combine the team-results and public-picks datasets to be one merged dataset.

```{r}
# merging the two datasets on a full join by their team names:

fullData <- full_join(team_results, public_picks, by = c("TEAM" = "TEAM"))


# converting the <char> percentages to doubles and removing the '%'

percent_columns <- c("R64.y", "R32.y", "S16.y", "E8.y", "F4.y", "F4PERCENT", "CHAMPPERCENT", "FINALS")
fullData[percent_columns] <- lapply(fullData[percent_columns], function(x) {
  as.numeric(gsub("%", "", x)) / 100
})

head(fullData)
```

We also generate a ratio of wins to losses
```{r}

fullData = fullData %>% 
  mutate(WLratio = W/L) 

head(fullData)

```


## RESULTS

### HYPOTHESIS 1 - SIROHI


### HYPOTHESIS 2 - CHAIRA


## DISCUSSION


### LIMITATIONS 



### FUTURE DIRECTION



## Data Appendix
